---
title: 'HW on Handling Large Datasets: Analyzing Baby Names in the US'
author: 'Karen Chik'
output: html_document
---


####Data

Acquire and analyze a real dataset on baby name popularity provided by the Social Security Administration.
The data can be downloaded in zip format from:
http://www.ssa.gov/oact/babynames/state/namesbystate.zip  (~22MB)


#### Read csv files in folder

Write a loop to read all files in folder
```{r}
setwd("/Users/Karen/Desktop/Github/namesbystate")
file_list<-list.files()
for(file in file_list){
if (!exists("dataset")){
    dataset <- read.table(file, header=FALSE, sep=",")}
   
  # if the merged dataset does exist, append to it
  if (exists("dataset")){
    temp_dataset <-read.table(file, header=FALSE, sep=",")
    dataset<-rbind(dataset, temp_dataset)
    rm(temp_dataset)}
}

colnames(dataset)<-c("State","Gender","Year","Name","Count")
```

#### Identify most popular baby names across both genders
```{r}
library(DT)
library(dplyr)
library(data.table)
dt1<-data.table(summarize(group_by(dataset,Name),sum(Count)))
idx<-which.max(dt1$`sum(Count)`)
dt1[idx]

```

#### Identify most gender ambiguous names in 2013 and 1945
Gender ambiguity metric= absolute value of male/female count - 1 per name, whichever name has ratio closest to 0 is the most gender ambiguous.

```{r}

yr13<-filter(dataset,Year==2013)

Gender_Ambg= function (yr_df){
  df=yr_df
dt2<-data.table(summarize(group_by(df,Gender,Name),sum(Count)))
dt2 <-  dt2 %>% group_by(Name) %>% filter(n()>1)
male<-filter(dt2,Gender=='M')
female<-filter(dt2,Gender=='F')
uniq_name<-data.table(unique(male$Name))
names(uniq_name)[1]<-"Name"

uniq_name$Ratio<-abs(((male$`sum(Count)`/female$`sum(Count)`)-1))
r_idx<-which(uniq_name$Ratio==min(uniq_name$Ratio))
uniq_name[r_idx,]
}
Gender_Ambg(yr13)

yr45<-filter(dataset,Year==1945)
Gender_Ambg(yr45)
```


#### Identify names with the largest percentage increase and decrease in popularity since 1980

```{r}
yr80<-filter(dataset,Year==1980)
yr15<-filter(dataset,Year==2015)

dt3<-data.table(summarize(group_by(yr80,Name),sum(Count)))
dt4<-data.table(summarize(group_by(yr15,Name),sum(Count)))
for (j in 2:length(dt3)) {
  Common_Names = merge(dt3,dt4,by="Name",all=TRUE)
}
colnames(Common_Names) <- c("Name","Count_1980","Count_2015")
Common_Names$Pct_Change<- ((Common_Names$Count_2015)-(Common_Names$Count_1980))/(Common_Names$Count_1980)
r_idx1<-which.max(Common_Names$Pct_Change)
Increase<-Common_Names[r_idx1]
r_idx2<-which.min(Common_Names$Pct_Change)
Decrease<-Common_Names[r_idx2]

Increase
Decrease
```


#### Identify names with the largest percentage increase and decrease in popularity across all years
```{r}

PopName=function(beg_yr,end_yr){
yrbeg<-filter(dataset,Year==beg_yr)
yrend<-filter(dataset,Year==2015)
yrbeg<-yrbeg%>%group_by(Name)%>%summarise(Count=sum(Count))
yrend<-yrend%>%group_by(Name)%>%summarise(Count=sum(Count))
Comm_Names = merge(yrbeg,yrend,by="Name",all=TRUE)
colnames(Comm_Names) <- c("Name","Count_Beg","Count_2015")
Comm_Names$Pct_Change<- ((Comm_Names$Count_2015)-(Comm_Names$Count_Beg))/(Comm_Names$Count_Beg)
Comm_Names[which.max(Comm_Names$Pct_Change),]
}

NPopName=function(beg_yr,end_yr){
yrbeg<-filter(dataset,Year==beg_yr)
yrend<-filter(dataset,Year==2015)
yrbeg<-yrbeg%>%group_by(Name)%>%summarise(Count=sum(Count))
yrend<-yrend%>%group_by(Name)%>%summarise(Count=sum(Count))
Comm_Names = merge(yrbeg,yrend,by="Name",all=TRUE)
colnames(Comm_Names) <- c("Name","Count_Beg","Count_2015")
Comm_Names$Pct_Change<- ((Comm_Names$Count_2015)-(Comm_Names$Count_Beg))/(Comm_Names$Count_Beg)
r_idx2<-which.min(Comm_Names$Pct_Change)
Comm_Names[r_idx2,]
}
Max_Names<-data.table(t(sapply(1910:2014, PopName,end_yr=2015)))
Max_Names$Year<-(seq(1910,2014))
Min_Names<-data.table(t(sapply(1910:2014, NPopName,end_yr=2015)))
Min_Names$Year<-(seq(1910,2014))

Max_Names[which.max(Max_Names$Pct_Change),] 
Min_Names[which.min(Min_Names$Pct_Change),]


```

#### Growth of baby names based on US presidential terms

```{r}
library(data.table)
library(xlsx)
library(dplyr)
library(tidyr)
require(reshape)

library(stringr)

prez<-read.xlsx("US presidents listed.xlsx",1)
prez<-data.table(prez)
prez<-prez%>%separate(President, into=paste0(c("Name",1:3),sep=" "))
colnames(prez)[1] <- "Year"
colnames(prez)[2]<-"Name"
prez<-subset(prez,select=c("Year","Name"))
baby_yr<-data.table(dataset%>%group_by(Name,Year)%>%summarise(Count=sum(Count)))
baby_yr = baby_yr %>% select(Year, Name, Count)
baby_yr$Year<-as.factor(baby_yr$Year)
prez$Year<-as.factor(prez$Year)
Comm_Names<-data.table(merge(baby_yr,prez,by=c("Year","Name")))

Pct_chg = function (x){
  (x-lag(x))/lag(x)
}
p_chg<-data.table(Comm_Names%>%group_by(Name)%>%mutate_each(funs(Pct_chg),Count))
p_chg<-na.omit(p_chg)
p_chg%>%group_by(Name)%>%summarise(Count=sum(Count))
```
Interesting findings:
27 babies were named (presumably) after President Obama in year 2009. 
The highest increase of baby names during a presidential term was William and Lyndon.
The highest decrease of baby names during a presidential term was Franklin.

